---
title: "SUMS Subgroup Analysis - Surrogate Measured Prior to Mortality Endpoint"
output: html_document
date: "2026-08-04"
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

#Load packages
library(gdata)
library(flextable)

#Load bootstrap function
setwd("../code/Bootstrap")
source("bootstrap_functions.R")

# Trials where surrogate was measured prior to the mortality endpoint:
# Breast: Gothenburg Breast, HIP, Russia WHO Breast, Stockholm Breast, UK Age
# Lung:   Czech study, ITALUNG, Lung Screening Study (LSS), LUSI
# Prostate: UK CAP, Norrkoping, PLCO (Prostate)

trials_prior <- c(
  "Gothenburg Breast",
  "HIP",
  "Russia WHO Breast",
  "Stockholm Breast",
  "UK Age",
  "Czech study",
  "ITALUNG",
  "Lung Screening Study (LSS)",
  "LUSI",
  "UK CAP",
  "Norrkoping",
  "PLCO (Prostate)"
)

#function to prepare data for analysis
prep_data <- function (finaldata_primary) {
  
  finaldata_primary <- finaldata_primary %>%
  # calculate weights (inverse variance 1/log(v))
  mutate(weight_var = 1/vi_m) %>%
  #only keep trials with available surrogate estimate
  filter(!is.na(yi_s)) %>%
  #count trials in analysis
  mutate(N_trials=n_distinct(trial_acro_sm))
  
  return(finaldata_primary)
}

#function to print results table
print_results <- function(results_df) {
  results_df %>%
    mutate(Slope_lb_ci = ifelse(Slope_lb_ci == "-99999", "-Inf", Slope_lb_ci)) %>%
    mutate(Slope_ub_ci = ifelse(Slope_ub_ci == "99999", "Inf", Slope_ub_ci)) %>%
    flextable() %>%
    set_header_labels(
      Ntrials = "N trials",
      Intercept_lb_ci = "Intercept 95% CI; lower",
      Intercept_ub_ci = "Intercept 95% CI; upper",
      Slope_lb_ci = "Slope 95% CI; lower",
      Slope_ub_ci = "Slope 95% CI; upper",
      Correlation_lb_ci = "Correlation 95% CI; lower",
      Correlation_ub_ci = "Correlation 95% CI; upper",
      Rsquared = "R-squared",
      Rsquared_lb_ci = "R-squared 95% CI; lower",
      Rsquared_ub_ci = "R-squared 95% CI; upper"
    ) %>%
    autofit()
}

```

# Subgroup: trials where surrogate measured prior to mortality endpoint

Trials included:

- **Breast**: Gothenburg Breast, HIP NY, Russia / WHO, Stockholm, UK Age
- **Lung**: Czech Study, ITALUNG, LSS, LUSI
- **Prostate**: CAP, Norrkoping, PLCO

---

## Surrogate 4: Absolute incidence of early stage cancer

```{r surrogate4, echo=FALSE, warning=FALSE}

# Create results table
results_surrogate_4 <- data.frame(Cancer=character(),
                                  Surrogate=character(),
                                  Analysis=character(),
                                  Ntrials=numeric(),
                                  Intercept=numeric(),
                                  Intercept_lb_ci=numeric(),
                                  Intercept_ub_ci=numeric(),
                                  Slope=numeric(),
                                  Slope_lb_ci=numeric(),
                                  Slope_ub_ci=numeric(),
                                  Correlation=numeric(),
                                  Correlation_lb_ci=numeric(),
                                  Correlation_ub_ci=numeric(),
                                  Rsquared=numeric(),
                                  Rsquared_lb_ci=numeric(),
                                  Rsquared_ub_ci=numeric()
                                  )

#Load cancer-specific data
setwd("../code/Estimates")
source("Lung.R")
lung_4 <- finaldata_primary_4 %>%
  mutate(cancer="Lung")

setwd("../code/Estimates")
source("Prostate.R")
prostate_4 <- finaldata_primary_4 %>%
  mutate(cancer="Prostate")

setwd("../code/Estimates")
source("Breast.R")
breast_4 <- finaldata_primary_4 %>%
  mutate(cancer="Breast") %>%
  mutate(trial_acro_sm=ifelse(trial_acro_sm=="Mumbai Breast Cervix","Mumbai Breast",trial_acro_sm))

gdata::keep(
  lung_4,
  prostate_4,
  breast_4,
  trials_prior,
  results_surrogate_4,
  analysis,
  prep_data,
  print_results,
  bootstrap_trial,
  sure = T
)

# Combine and restrict to trials where surrogate was measured prior to mortality
all_data_4 <- bind_rows(lung_4, prostate_4, breast_4) %>%
  filter(trial_acro_sm %in% trials_prior)

all_data_4 <- prep_data(all_data_4)

#Analysis (see function in Allcancer_functions.R)
results_surrogate_4 <- results_surrogate_4 %>%
  add_row(analysis(all_data_4, "All", "4. Absolute incidence of early stage cancer", "Surrogate prior to mortality", bootstrap_trial))

# Print results table
print_results(results_surrogate_4)

# Bubble plot
ggplot(all_data_4, aes(x = exp(yi_s), y = exp(yi_m), label=trial_acro_sm, weight=weight_var)) +
  geom_point(aes(size = weight_var, colour=cancer), shape = 16) +
  geom_abline(data=results_surrogate_4, aes(intercept = Intercept[Cancer=="All"], slope = Slope[Cancer=="All"])) +
  geom_hline(yintercept = 1, linetype=3)+
  geom_vline(xintercept = 1, linetype=3)+
  scale_x_continuous(trans="log", limits = c(0.3,6), breaks=c(0.5,1,2,5)) +
  scale_y_continuous(trans="log", limits = c(0.3,6), breaks=c(0.5,1,2,5)) +
  labs(x="RR early stage incidence", y="RR Mortality", colour="Cancer type")+
  geom_text(hjust=-0.05, vjust=-0.5)+
  theme_bw() +
  theme(plot.caption=element_text(hjust=0), aspect.ratio = 1) +
  guides(size = FALSE)

```

---

## Surrogate 5: Proportion of target cancers that are screen-detected

```{r surrogate5, echo=FALSE, warning=FALSE}

# Create results table
results_surrogate_5 <- data.frame(Cancer=character(),
                                  Surrogate=character(),
                                  Analysis=character(),
                                  Ntrials=numeric(),
                                  Intercept=numeric(),
                                  Intercept_lb_ci=numeric(),
                                  Intercept_ub_ci=numeric(),
                                  Slope=numeric(),
                                  Slope_lb_ci=numeric(),
                                  Slope_ub_ci=numeric(),
                                  Correlation=numeric(),
                                  Correlation_lb_ci=numeric(),
                                  Correlation_ub_ci=numeric(),
                                  Rsquared=numeric(),
                                  Rsquared_lb_ci=numeric(),
                                  Rsquared_ub_ci=numeric()
                                  )

#Load cancer-specific data
setwd("../code/Estimates")
source("Lung.R")
lung_5 <- finaldata_primary_5 %>%
  mutate(cancer="Lung")

setwd("../code/Estimates")
source("Prostate.R")
prostate_5 <- finaldata_primary_5 %>%
  mutate(cancer="Prostate")

setwd("../code/Estimates")
source("Breast.R")
breast_5 <- finaldata_primary_5 %>%
  mutate(cancer="Breast") %>%
  mutate(trial_acro_sm=ifelse(trial_acro_sm=="Mumbai Breast Cervix","Mumbai Breast",trial_acro_sm))

gdata::keep(
  lung_5,
  prostate_5,
  breast_5,
  trials_prior,
  results_surrogate_5,
  analysis,
  prep_data,
  print_results,
  bootstrap_trial,
  sure = T
)

# Combine and restrict to trials where surrogate was measured prior to mortality
all_data_5 <- bind_rows(lung_5, prostate_5, breast_5) %>%
  filter(trial_acro_sm %in% trials_prior)

all_data_5 <- prep_data(all_data_5)

#Analysis (see function in Allcancer_functions.R)
results_surrogate_5 <- results_surrogate_5 %>%
  add_row(analysis(all_data_5, "All", "5. Proportion of target cancers that are screen-detected", "Surrogate prior to mortality", bootstrap_trial))

# Print results table
print_results(results_surrogate_5)

# Bubble plot
ggplot(all_data_5, aes(x = yi_s, y = exp(yi_m), label=trial_acro_sm, weight=weight_var)) +
  geom_point(aes(size = weight_var, colour=cancer), shape = 16) +
  geom_abline(data=results_surrogate_5, aes(intercept = Intercept[Cancer=="All"], slope = Slope[Cancer=="All"])) +
  geom_hline(yintercept = 1, linetype=3)+
  scale_y_continuous(trans="log", limits = c(0.3,6), breaks=c(0.5,1,2,5)) +
  scale_x_continuous(limits=c(0,1))+
  labs(x="% of target cancers that are screen-detected", y="RR Mortality", colour="Cancer type")+
  geom_text(hjust=-0.05, vjust=-0.5)+
  theme_bw() +
  theme(plot.caption=element_text(hjust=0), aspect.ratio = 1) +
  guides(size = FALSE)

```

---

## Surrogate 6: Proportion of high-grade target cancers that are screen-detected

```{r surrogate6, echo=FALSE, warning=FALSE}

# Create results table
results_surrogate_6 <- data.frame(Cancer=character(),
                                  Surrogate=character(),
                                  Analysis=character(),
                                  Ntrials=numeric(),
                                  Intercept=numeric(),
                                  Intercept_lb_ci=numeric(),
                                  Intercept_ub_ci=numeric(),
                                  Slope=numeric(),
                                  Slope_lb_ci=numeric(),
                                  Slope_ub_ci=numeric(),
                                  Correlation=numeric(),
                                  Correlation_lb_ci=numeric(),
                                  Correlation_ub_ci=numeric(),
                                  Rsquared=numeric(),
                                  Rsquared_lb_ci=numeric(),
                                  Rsquared_ub_ci=numeric()
                                  )

#Load cancer-specific data
setwd("../code/Estimates")
source("Prostate.R")
prostate_6 <- finaldata_primary_6 %>%
  mutate(cancer="Prostate")

setwd("../code/Estimates")
source("Breast.R")
breast_6 <- finaldata_primary_6 %>%
  mutate(cancer="Breast") %>%
  mutate(trial_acro_sm=ifelse(trial_acro_sm=="Mumbai Breast Cervix","Mumbai Breast",trial_acro_sm))

gdata::keep(
  prostate_6,
  breast_6,
  trials_prior,
  results_surrogate_6,
  analysis,
  prep_data,
  print_results,
  bootstrap_trial,
  sure = T
)

# Combine and restrict to trials where surrogate was measured prior to mortality
all_data_6 <- bind_rows(prostate_6, breast_6) %>%
  filter(trial_acro_sm %in% trials_prior)

all_data_6 <- prep_data(all_data_6)

#Analysis (see function in Allcancer_functions.R)
results_surrogate_6 <- results_surrogate_6 %>%
  add_row(analysis(all_data_6, "All", "6. Proportion of high-grade target cancers that are screen-detected", "Surrogate prior to mortality", bootstrap_trial))

# Print results table
print_results(results_surrogate_6)

# Bubble plot
ggplot(all_data_6, aes(x = yi_s, y = exp(yi_m), label=trial_acro_sm, weight=weight_var)) +
  geom_point(aes(size = weight_var, colour=cancer), shape = 16) +
  geom_abline(data=results_surrogate_6, aes(intercept = Intercept[Cancer=="All"], slope = Slope[Cancer=="All"])) +
  geom_hline(yintercept = 1, linetype=3)+
  scale_y_continuous(trans="log", limits = c(0.3,2), breaks=c(0.5,1,1.5,2)) +
  scale_x_continuous(limits=c(0,1))+
  labs(x="% of high-grade target cancers that are screen-detected", y="RR Mortality", colour="Cancer type")+
  geom_text(hjust=-0.05, vjust=-0.5)+
  theme_bw() +
  theme(plot.caption=element_text(hjust=0), aspect.ratio = 1) +
  guides(size = FALSE)

```

---

## Surrogate 7: Diagnostic yield of screening

```{r surrogate7, echo=FALSE, warning=FALSE}

# Create results table
results_surrogate_7 <- data.frame(Cancer=character(),
                                  Surrogate=character(),
                                  Analysis=character(),
                                  Ntrials=numeric(),
                                  Intercept=numeric(),
                                  Intercept_lb_ci=numeric(),
                                  Intercept_ub_ci=numeric(),
                                  Slope=numeric(),
                                  Slope_lb_ci=numeric(),
                                  Slope_ub_ci=numeric(),
                                  Correlation=numeric(),
                                  Correlation_lb_ci=numeric(),
                                  Correlation_ub_ci=numeric(),
                                  Rsquared=numeric(),
                                  Rsquared_lb_ci=numeric(),
                                  Rsquared_ub_ci=numeric()
                                  )

#Load cancer-specific data
setwd("../code/Estimates")
source("Lung.R")
lung_7 <- finaldata_primary_7 %>%
  #exclude trials with no unscreened control arm (as per primary analysis for surrogate 7)
  filter(comparator_other_screening=="Not applicable") %>%
  mutate(cancer="Lung")

setwd("../code/Estimates")
source("Prostate.R")
prostate_7 <- finaldata_primary_7 %>%
  filter(comparator_other_screening=="Not applicable") %>%
  mutate(cancer="Prostate")

setwd("../code/Estimates")
source("Breast.R")
breast_7 <- finaldata_primary_7 %>%
  filter(comparator_other_screening=="Not applicable") %>%
  mutate(cancer="Breast") %>%
  mutate(trial_acro_sm=ifelse(trial_acro_sm=="Mumbai Breast Cervix","Mumbai Breast",trial_acro_sm))

gdata::keep(
  lung_7,
  prostate_7,
  breast_7,
  trials_prior,
  results_surrogate_7,
  analysis,
  prep_data,
  print_results,
  bootstrap_trial,
  sure = T
)

# Combine and restrict to trials where surrogate was measured prior to mortality
all_data_7 <- bind_rows(lung_7, prostate_7, breast_7) %>%
  filter(trial_acro_sm %in% trials_prior)

all_data_7 <- prep_data(all_data_7)

#Analysis (see function in Allcancer_functions.R)
results_surrogate_7 <- results_surrogate_7 %>%
  add_row(analysis(all_data_7, "All", "7. Diagnostic yield of screening", "Surrogate prior to mortality", bootstrap_trial))

# Print results table
print_results(results_surrogate_7)

# Bubble plot
ggplot(all_data_7, aes(x = yi_s, y = exp(yi_m), label=trial_acro_sm, weight=weight_var)) +
  geom_point(aes(size = weight_var, colour=cancer), shape = 16) +
  geom_abline(data=results_surrogate_7, aes(intercept = Intercept[Cancer=="All"], slope = Slope[Cancer=="All"])) +
  geom_hline(yintercept = 1, linetype=3)+
  scale_y_continuous(trans="log", limits = c(0.1,6), breaks=c(0.5,1,1.5,2,3,5)) +
  labs(x="Diagnostic yield of screening (%)", y="RR Mortality", colour="Cancer type")+
  geom_text(hjust=-0.05, vjust=-0.5)+
  theme_bw() +
  theme(plot.caption=element_text(hjust=0), aspect.ratio = 1) +
  guides(size = FALSE)

```
